Modeling the impact of IT- and AI-related emissions in a university setting

### 2025-08-04 doc/paper.tex
Finished results draft

For desktops, most estimates are in the range of 250 - 400 \kgcoe. The most detailed data comes from Dell's Product Carbon Footprints (\cite{dellpcf}), which the company publishes for each of their products. A sample of 15 desktops has mean supply chain emissions of 289 \kgcoe\ with $\sigma = 80$ kg. We model desktop embodied emissions at 290 \kgcoe\ / device.

@@

\begin{itemize}
  \item Computer monitors: 340 \kgcoe\ (\cite{teehan2013}, \cite{dellpcf}).
  \item Conference room displays: 700 \kgcoe\ (scaling monitor by $(40''/27'')^2$). We use this as a proxy for all audiovisual equipment (lecture room projectors, etc.). Note that the numbers are sufficiently low that errors in this estimate do not significantly affect the overall result.
  \item Printer/copier stations: 1100 \kgcoe\ (\cite{ecoinvent}).
  \item Network equipment: We did not find reliable lifecycle assessments for routers and switches, and have little information about UZH network architecture. The vast majority ($80-95\%$) of lifecycle emissions of network gear stem from usage (\cite{cisco2024}, \cite{jacob2023}). For now, we model the embodied emissions of network equipment like we do its operational emissions, namely as a 10\% overhead on the number of devices (laptops, desktops, servers).
\end{itemize}

@@

Maybe the most striking result is that, when we consider the entire population of campus devices, and not only those owned and managed by the university, overall emissions are absolutely dominated by the manufacturing and disposal of end-user devices. See Figure~\ref{fig:1}.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\textwidth]{fig-baseline.png}
  \caption{Baseline results}
  \label{fig:1}
\end{figure}

@@

Hypothetically moving all UZH server workloads from on-premise data centers to a commercial cloud provider eliminates about 170 \tcoe\ of cooling overhead, and replaces 415 \tcoe\ of total server emissions with 380 \tcoe\ of cloud emissions.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\textwidth]{fig-cloud.png}
  \caption{Migration to cloud}
  \label{fig:1}
\end{figure}

@@

@@

>>>>

For desktops, most estimates are in the range of 250 - 400 \kgcoe. The most detailed data comes from Dell's Product Carbon Footprints~\cite{dellpcf}, which the company publishes for each of their products. A sample of 15 desktops has mean supply chain emissions of 289 \kgcoe\ with $\sigma = 80$ kg. We model desktop embodied emissions at 290 \kgcoe\ / device.

@@

\begin{itemize}
  \item Computer monitors: 340 \kgcoe~\cite{dellpcf,lovehagen2023,teehan2013}.
  \item Conference room displays: 700 \kgcoe\ (scaling monitor by $(40''/27'')^2$). We use this as a proxy for all audiovisual equipment (lecture room projectors, etc.). Note that the numbers are sufficiently low that errors in this estimate do not significantly affect the overall result.
  \item Printer/copier stations: 1100 \kgcoe~\cite{ecoinvent}.
  \item Network equipment: We did not find reliable lifecycle assessments for routers and switches, and have little information about UZH network architecture. The vast majority ($80-95\%$) of lifecycle emissions of network gear stem from usage~\cite{cisco2024,jacob2023}. For now, we model the embodied emissions of network equipment like we do its operational emissions, namely as a 10\% overhead on the number of devices (laptops, desktops, servers).
\end{itemize}

@@

Maybe the most striking result is that, when we consider the entire population of campus devices, and not only those owned and managed by the university, overall emissions are absolutely dominated by the manufacturing and disposal of end-user devices. See Figure~\ref{fig:baseline}.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\textwidth]{fig-baseline.png}
  \caption{Baseline results}
  \label{fig:baseline}
\end{figure}

@@

Hypothetically moving all UZH server workloads from on-premise data centers to a commercial cloud provider eliminates about 170 \tcoe\ of cooling overhead, and replaces 415 \tcoe\ of total server emissions with 380 \tcoe\ of cloud emissions. See Figure~\ref{fig:cloud}.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\textwidth]{fig-cloud.png}
  \caption{Migration to cloud}
  \label{fig:cloud}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\textwidth]{fig-llm-growth.png}
  \caption{LLM growth projection}
  \label{fig:llm:growth}
\end{figure}

AI adoption is growing rapidly, driving demand for data centers and potentially stressing electrical infrastructure in coming years~\cite{lbl:dc:2024,iea:ai:energy}. However, for an institution of roughly constant size such as UZH, and where IT infrastructure is already well developed and heavily used, emissions from generative AI most likely represent a small addition to overall emissions. In our model, using the default parameters, they comprise just 1\% of total emissions.

Moreover, under even aggressive growth projections (50\% YoY usage growth, 50\% YoY training overhead growth, a moderate 20\% YoY improvement in inference efficiency), that overhead remains under 3\% at the end of 5 years. See Figure~\ref{fig:llm:growth}.

These results are consistent with other analyses that attempt to put LLM use in the context of other computer activity~\cite{epoch2025howmuchenergydoeschatgptuse,ritchie:chatgpt}.

@@

The results above are robust under remarkably different assumptions.

For example, if we (1) massively reduce the set of end-user devices in scope to just those owned and managed by the university (about 10,000 laptops and monitors, 2,400 mobile phones), (2) triple the LLM per-query energy use to 3 Wh, and (3) assume that each of the 38,000 people at UZH uses a GPT-4o-class LLM 40 times per day, then emissions from AI still remain around 10\% of total (Figure~\ref{fig:sens1}).

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\textwidth]{fig-sens1.png}
  \caption{A scenario with only university-managed end-user devices}
  \label{fig:sens1}
\end{figure}

Similarly, if we cut our estimate for the embodied emissions of every type of device to the lowest value found in the literature (as opposed to a more representative middle-of-the-road value), but keep all other parameters at their default values, embodied emissions {\em still} dominate and LLM emissions continue to be a small fraction of the total, as seen in Figure~\ref{fig:sens2}.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\textwidth]{fig-sens2.png}
  \caption{A scenario with minimal embodied emissions}
  \label{fig:sens2}
\end{figure}

Only if we combine the ``low embodied emissions'' scenario with the ``pessimal LLM'' scenario does generative AI become a major factor, contributing 18\% of total emissions.
